Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 conclusions.tex
lloid_paper: Many small fixes and fill-ins
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@906 ccdd05d5-56c1-461a-bf98-85f6876523b4

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence introducing an up-front latency of 125~ms.  Data
%
\editorial{This is redundant: 1/(16~Hz) and 125~ms are just reciprocals of each other.}%
%
aggregation from the observatories will travel over various networks, each
capable of high bandwith but perhaps only modest latency.  This could amount to
a similar latency of $\sim$ 100 ms.  Lastly, unless a realtime infrastructure is
adopted post data acquisition, it is likely that there will be an inherent
latency introduced by such infrastructure.  We have shown a prototype
implementation using \gstlal\ that is capable of $\sim$1 s latency. In our
opinion, significant work would have to be done in order to improve upon this
number. However, it should be considered for third generation detector design.
For example, a tighter integration of analysis and data acquisition would be
beneficial.

\editorial{The way this is worded, it sounds like a big omission.}%
We have omitted discussion of source localization. Localization is known to
be poor for signals of low \SNR~\cite{Fairhurst2009}.
However, one should not immediately dismiss the practical usefulness of a
poorly localized source. Even with poor localization, it should be possible to
begin downselecting what observatories could view a potential signal and for
such observatories to begin any necessary prerequisite activities. In future
works we wil explore more rigorously the pointing prospects with realistic
simulations using the infrastructure and techniques described in this work.

>>>>

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence introducing an up-front latency of 125~ms.  Data
%
\editorial{This is redundant: 1/(16~Hz) and 125~ms are just reciprocals of each other.}%
%
aggregation from the observatories will travel over various networks, each
capable of high bandwith but perhaps only modest latency.  This could amount to
a similar latency of $\sim$100 ms.  Lastly, unless a realtime infrastructure is
adopted post data acquisition, it is likely that there will be an inherent
latency introduced by such infrastructure.  We have shown a prototype
implementation using \gstlal\ that is capable of $\sim$1 s latency. In our
opinion, significant work would have to be done in order to improve upon this
number. However, it should be considered for third-generation detector design.
For example, a tighter integration of analysis and data acquisition would be
beneficial.

\editorial{The way this is worded, it sounds like a big omission.}%
We have omitted discussion of source sky localization. Localization is known to
be poor for signals of low \SNR~\cite{Fairhurst2009}.
However, one should not immediately dismiss the practical usefulness of a
poorly localized source. Even with poor localization, it should be possible to
begin downselecting what observatories could view a potential signal and for
such observatories to begin any necessary prerequisite activities. In future
works we wil explore more rigorously the pointing prospects with realistic
simulations using the infrastructure and techniques described in this work.
\editorial{nvf: It's a bit pie-in-the-sky, but we could also mention reconfiguring
the signal recycling mirror to optimize SNR at merger. There's a lot of science there.}

### 2011-06-03 figures/diagram.tex
Corrected error in second delay
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@821 ccdd05d5-56c1-461a-bf98-85f6876523b4
lloid_paper: Many small fixes and fill-ins
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@906 ccdd05d5-56c1-461a-bf98-85f6876523b4

	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay0) {\tiny$z^{-t^1 f^0}$};
	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay1) {\tiny$z^{-t^2 f^1}$};
	\draw[->] (in0) -- (in1);
	\draw (in1) -- (invdots);
	\draw[->] (invdots) -- (in2);
	\draw[->] (x) -- (in0);

	\draw[decorate, decoration={brace, amplitude=4pt}] ($(in)+(-4mm,0)$) -- ($(in)+(4mm,0)$) node[midway,above=2mm,align=center] {Decimation \\ of input};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(intee)+(-2mm,0)$) -- ($(u)+(2mm,0)$) node[midway,above=2mm,align=center] {Orthogonal \\ FIR filters};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(utee)+(-2mm,0)$) -- ($(v)+(2mm,0)$) node[midway,above=2mm,align=center] {Reconstruction \\ matrices};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(add0)+(-2mm,0)$) -- ($(add3)+(2mm,0)$) node[midway,above=2mm,align=center] {Interpolation and \\ SNR accumulation};
\end{tikzpicture}

>>>>

	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay0) {\tiny$z^{-t^1 f^0}$};
	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay1) {\tiny$z^{-(t^2-t^1) f^1}$};
	\draw[->] (in0) -- (in1);
	\draw (in1) -- (invdots);
	\draw[->] (invdots) -- (in2);
	\draw[->] (x) -- (in0);

	\draw[decorate, decoration={brace, amplitude=4pt}] ($(in)+(-4mm,0)$) -- ($(in)+(4mm,0)$) node[midway,above=2mm,align=center] {Decimation \\ of input};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(intee)+(-2mm,0)$) -- ($(u)+(2mm,0)$) node[midway,above=2mm,align=center] {Orthogonal \\ \textsc{fir} filters};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(utee)+(-2mm,0)$) -- ($(v)+(2mm,0)$) node[midway,above=2mm,align=center] {Reconstruction \\ matrices};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(add0)+(-2mm,0)$) -- ($(add3)+(2mm,0)$) node[midway,above=2mm,align=center] {Interpolation and \\ \textsc{snr} accumulation};
\end{tikzpicture}

### 2011-06-03 implementation.tex
Added missing period
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@904 ccdd05d5-56c1-461a-bf98-85f6876523b4
lloid_paper: Many small fixes and fill-ins
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@906 ccdd05d5-56c1-461a-bf98-85f6876523b4

\begin{table}
\begin{indented}
\caption{Filter design for these 657 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  The \textsc{svd} tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$}
\item[]\begin{tabular}{rr@{,}lc*{6}{r}}
\br
\multicolumn{4}{c}{} &\centre{6}{$\log_{10}$ (1 - \textsc{svd} tolerance)} \\
\ns
\multicolumn{4}{c}{} &\crule{6} \\
$f^s$ (Hz) & $(t^{s+1}$&$t^s]$ (s) & $N^s$ & -1 & -2 & -3 & -4 & -5 & -6 \\
\mr
4096 & (0.5&0] & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & (4.5&0.5] & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & (12.5&4.5] & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & (76.5&12.5] & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & (140.5&76.5] & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & (268.5&140.5] & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & (396.5&268.5] & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & (460.5&396.5] & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & (588.5&460.5] & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & (844.5&588.5] & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & (1100.5&844.5] & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

The \lloid\ algorithm could be used in a true sample-in-sample-out real-time
system.  However, such a system would likely require integration directly into
the data acquisition and storage system of the gravitational wave
observatories.  A slightly more modest goal is to leverage existing low
latency, but not real-time, signal processing infrastructure in order to
implement the \lloid\ algorithm.  For the near-term this should be a viable 
solution for ultra-low latency searches with $\sim 1s$ intrinsic latency.

We have implemented a prototype of the low latency filtering stage using an
open source signal processing environment called \gstreamer\ \cite{gstreamer}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of gravitational
wave detector data to audio data it is not surprising that \gstreamer\ is
useful for our purpose. \gstreamer\ also provides some useful stock signal
processing elements such as resamplers and filters.  We have extended the
\gstreamer\ framework by developing a library called \gstlal{}~\cite{gstlal}
that provides elements for gravitational wave data analysis.  Figure
\ref{fig:pipeline} describes schematically how we implement the \lloid\ algorithm
using \gstlal\ and \gstreamer\ components.
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of LLOID pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for FIR filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}

The signal flow diagram drawn in figure \ref{fig:pipeline} contains several
distinct stages and parallel branches.  The stages are decimation, filtering
the data against the orthogonal templates, reconstruction of the original template basis, interpolation
and accumulation of the early warning and final {\SNR}.  We outline the stages
below.

@@

First, the sample rate of the whitened detector data is reduced to successively
lower sample rates by decimation.  Decimation involves applying an anti-aliasing
filter to the data, and then down-sampling by deleting samples.  We use a
192-tap \fir\ decimator provided by {\tt Gstreamer}'s {\tt audioresample}
element.  The detector data is provided at every power-of-two sample rate
required by the template time slices described in \eqref{eq:time-slices}.  These
parallel decimated data streams are fed into parallel \fir\ filter banks in the
next stage.

@@

The \fir\ filter banks are implemented using a \gstlal\ element called {\tt
lal\_firbank}, which produces N channels of filter output from an NxM matrix of
\fir\ filter coefficients.  This element is used in parallel branches in the
pipeline to implement the \SVD\ basis filters in each time
slice.  Rather than implement the time sliced templates as zero padded \fir\
filters as described in \eqref{eq:time-slices} we instead implement them as shorter
filters that contain only the nonzero samples.  Adding the appropriate time
offset to the filter output later in the pipeline makes up for the lack of
explicit zero padding.  The orthogonal filter outputs must next be reconstructed
into the output of the underlying time sliced templates.

@@

In order to form the early-warning \SNR\ from each time slice, we hav eto add the partial 
\SNR\ to the early-warning \SNR\ from the subsequent time slices.  If the next time slice 
does not have the same sample rate, its output must first be interpolated.  This is done
using the same \gstreamer\ element as was used for decimation, {\tt
audioresample}.

>>>>

\begin{table}
\begin{indented}
\caption{Filter design for these 657 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  The \SVD{} tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\item[]\begin{tabular}{rr@{,\,}lc*{6}{r}}
\br
\multicolumn{4}{c}{} &\centre{6}{$\log_{10}$ (1 $-$ \textsc{svd} tolerance)} \\
\ns
\multicolumn{4}{c}{} &\crule{6} \\
$f^s$ (Hz) & $(t^{s+1}$&$t^s]$ (s) & $N^s$ & $-1$ & $-2$ & $-3$ & $-4$ & $-5$ & $-6$ \\
\mr
4096 & (0.5&0] & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & (4.5&0.5] & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & (12.5&4.5] & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & (76.5&12.5] & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & (140.5&76.5] & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & (268.5&140.5] & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & (396.5&268.5] & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & (460.5&396.5] & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & (588.5&460.5] & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & (844.5&588.5] & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & (1100.5&844.5] & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

The \lloid\ algorithm could be used in a true sample-in-sample-out real-time
system.  However, such a system would likely require integration directly into
the data acquisition and storage system of the gravitational-wave
observatories.  A slightly more modest goal is to leverage existing low
latency, but not real-time, signal processing infrastructure in order to
implement the \lloid\ algorithm.  For the near-term this should be a viable 
solution for ultra-low latency searches with $\sim 1s$ intrinsic latency.

We have implemented a prototype of the low-latency filtering stage using an
open-source signal processing environment called \gstreamer\ \cite{gstreamer}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of
gravitational- wave detector data to audio data it is not surprising that
\gstreamer\ is
useful for our purpose. \gstreamer\ also provides some useful stock signal
processing elements such as resamplers and filters.  We have extended the
\gstreamer\ framework by developing a library called \gstlal{}~\cite{gstlal}
that provides elements for gravitational-wave data-analysis.  Figure
\ref{fig:pipeline} describes schematically how we implement the \lloid\ algorithm
using \gstlal\ and \gstreamer\ components.
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}

The signal flow diagram drawn in figure \ref{fig:pipeline} contains several
distinct stages and parallel branches.  The stages are decimation, filtering
the data against the orthogonal templates, reconstruction of the original template basis, interpolation
and accumulation of the early-warning and final {\SNR}.  We outline the stages
below.

@@

First, the sample rate of the whitened detector data is reduced to successively
lower sample rates by decimation.  Decimation involves applying an anti-aliasing
filter to the data, and then down-sampling by deleting samples.  We use a
192-tap \fir\ decimator provided by \gstreamer{}'s {\tt audioresample}
element.  The detector data is provided at every power-of-two sample rate
required by the template time slices described in \eqref{eq:time-slices}.  These
parallel decimated data streams are fed into parallel \fir\ filter banks in the
next stage.

@@

The \fir\ filter banks are implemented using a \gstlal\ element called {\tt
lal\_firbank}, which produces $N$ channels of filter output from an $N\times M$ matrix of
\fir\ filter coefficients.  This element is used in parallel branches in the
pipeline to implement the \SVD\ basis filters in each time
slice.  Rather than implement the time-sliced templates as zero-padded \fir\
filters as described in \eqref{eq:time-slices} we instead implement them as shorter
filters that contain only the nonzero samples.  Adding the appropriate time
offset to the filter output later in the pipeline makes up for the lack of
explicit zero-padding.  The orthogonal filter outputs must next be reconstructed
into the output of the underlying time-sliced templates.

@@

In order to form the early-warning \SNR\ from each time slice, we have to add
the partial \SNR\ to the early-warning \SNR\ from the subsequent time slices.
If the next time slice does not have the same sample rate, its output must
first be interpolated.  This is done using the same \gstreamer\ element as was
used for decimation, {\tt audioresample}.

### 2011-06-03 introduction.tex
lloid_paper: Many small fixes and fill-ins
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@906 ccdd05d5-56c1-461a-bf98-85f6876523b4

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\textsc{ligo} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current gravitational wave search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm for
\textsc{fft} convolution, currently the preferred method in gravitational wave
searches.  The most obvious example of the latter is the time domain
(\textsc{td}) convolution, which has no latency at all.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

>>>>

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current gravitational wave search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm for
\fft{} convolution, currently the preferred method in gravitational wave
searches.  The most obvious example of the latter is the time domain
(\textsc{td}) convolution, which has no latency at all.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.
